\bf Teaching analysis reproducibility with accessibility: Enriching secondary documentation to support skill level diversity of users.

### 2022-02-09 jsdse/Manuscript.tex
reviews: more mega lab meeting comments

%% I think I have three different points here:
%% internalizing knowledge is made easier with active learning practices: learn by doing!
%% literate programming allows learning by doing AND at the same time provides a story that connects the acquired knowledge to a real worl problem.
%% By definition, all languages are abstract. We learn them because we grow up with them and they have a direct application for our lives.
%% Programming languages are no different, however, we start learning those later in life and we do not use them as regularly, making it harder for us to learn them and use them.
Pedagogical research shows that active learning practices are one the most effective
ways to take on abstract subjects \citep{freeman2014active}.
Programming computer languages are quite abstract and learning them can be greatly
enhanced by applying an active learning strategy such as a linking its usage to
a ``real world'' or ``human'' application \citep{felder2009active}.

A story-like narrative that links pieces of code together and invites learners
to try the code can greatly support learners to remember what they are doing and
why they are doing it.
This ``literate programming'' paradigm \citep{knuth1984literate, fritzson2002mathmodelica}
makes code more approachable, as it integrates narratives with computer code in
the same document, supporting learners in actively following, remembering and understanding
the code usage \citep{piccolo2016tools}.
Documents developed with ``literate programming'' can be made more accessible by
choosing narratives that are relatable to a more general audience.

@@

We identify ways to evaluate inputs to know if they will produce an error, and design
alternatives on what to do when faced with an error or warning, and demonstrate
these alternatives.
One of the most essential skills in programming is interpreting and moving forward
from errors.
Many finely honed tutorials do not trigger errors, which precludes helping students
to develop the tools to understand and address errors when they do encounter them,
as they inevitably will.
On our tutorial, we focus on explaining the meaning and downstream of warnings and errors, and
 showcase ways to detect them before they are triggered (i.e., before using an input
  that would elicit a warning or error). This has two pedagogical benefits:
1) it provides users/students with the means to troubleshoot their own warnings and errors, and
2) it allows users/students to understand with more depth what the function is doing.

@@

Besides avoiding formal language, and incorporating elements of pop culture, such as picture
character icons known as ``emojis``, to make the language more familiar to a
broader target audience (see Figure \ref{fig:first}), we made an effort to specifically
complement the primary documentation by identifying
computational concepts that were assumed or were not explained in depth.
We vetted the tutorials with an audience on workshops as well as individual user.
We choose examples that are charismatic for the audience.
For example, when we presented the tutorial for a team specialized in Amphibians,
we tailored the examples using frogs and their allies.

@@

We published the tutorials on a public, free license, free of cost, and free for
use and reuse repository and persistent website \citep{RopentreeTutorials, RopentreeTutorialsWebsite}.
The tutorial is available for the users to go back to it any time they need it,
and to be passed on to other users (Figure \ref{fig:second}).

@@

Following the carpentries, we created a main version of the tutorial that
is updated. Versions presented on workshops are a copy from the original repository,
and represent a stable and temporal snapshot of the functions and workflows presented
in the tutorial.

@@

@@

Some universities have been incorporating the subject in their classes (see
\cite{uwlibraries2022, nigms2022}).
The focus of this resources has been for students to develop skills to document their work.
The principles identified and outlined here can be used to set learning goals and
outcomes on new reproducibility syllabus.

We have received emails from senior researchers thanking us for this materials,
 and students have been able to engage using the packages with less hep from the PIs.

The principles to create tutorials described here facilitate adoption of software
 and analysis workflows among researchers at different academic levels, from undergrads
  to established researchers.
It will also help closing the gap between students that had access to computational
 resources (and computational training) from an early age and students that did not.
  Late access to computational resources and training can occur due to lack of
   economic resources, often occurring in households from underrepresented communities
    and minorities.
It can also be due to gender-biased parental and community pressures,
 in which male individuals are more often encouraged to perform activities related to computers,
  while female individuals are discouraged.
% How to balance software acceptance VS. adoption?
These principles can be used to aide not only in improving reproducibility practices,
 but also software adoption in the natural sciences.
%% Discuss: why address accessibility and not other aspects of reproducibility?

Making accessible reproducible workflows has several advantages:
save explanation/training time when analyses are run again by students and collaborators.
save research time for yourself when analyses are run again with more data, a different dataset, a different organism or biological model.
scientific efforts can build off of each other

>>>>

%% I think I have three different points here:
%% internalizing knowledge is made easier with active learning practices: learn by doing!
%% literate programming allows learning by doing AND at the same time provides a story that connects the acquired knowledge to a real worl problem.
%% By definition, all languages are abstract. We learn them because we grow up with them and they have a direct application for our lives.
%% Programming languages are no different, however, we start learning those later in life and we do not use them as regularly, making it harder for us to learn them and use them.
Pedagogical research shows that active learning practices are one the most effective
ways to take on abstract subjects \citep{freeman2014active}.
Programming computer languages are quite abstract and learning them can be greatly
enhanced by applying an active learning strategy such as linking its usage to
a ``real world'' or ``human'' application \citep{felder2009active}.

A story-like narrative that links pieces of code together and invites learners
to try the code can lead learners to remember what they are doing and
why they are doing it.
This ``literate programming'' paradigm \citep{knuth1984literate, fritzson2002mathmodelica}
makes code more approachable, as it integrates narratives with computer code in
the same document, supporting learners in actively following, remembering and understanding
the code usage \citep{piccolo2016tools}.
Documents developed with ``literate programming'' can be made more accessible by
choosing narratives that are relatable to a more general audience.

@@

We identify ways to evaluate inputs to know if they will produce an error, and design
alternatives on what to do when faced with an error or warning, and demonstrate
these alternatives.
One of the most essential skills in programming is interpreting and moving forward
from errors.
Many finely honed tutorials do not trigger errors, which precludes helping students
to develop the tools to understand and address errors when they do encounter them,
as they inevitably will.
In our tutorial, we focus on explaining the meaning and downstream of warnings and errors, and
 showcase ways to detect them before they are triggered (i.e., before using an input
  that would elicit a warning or error). This has two pedagogical benefits:
1) it provides users/students with the means to troubleshoot their own warnings and errors, and
2) it allows users/students to understand with more depth what the function is doing.

@@

Besides avoiding formal language, and incorporating elements of pop culture, such as picture
character icons known as ``emojis``, to make the language more familiar to a
broader target audience (see Figure \ref{fig:first}), we made an effort to specifically
complement the primary documentation by identifying
computational concepts that were assumed or were not explained in depth.
We vetted the tutorials through feedback from workshop participants as well as individual users.
We choose examples that are charismatic for the audience.
For example, when we presented the tutorial for a team specialized in Amphibians,
we tailored the examples using frogs and their allies.

@@

We published the tutorials on a public, free license, free of cost, and free for
use and reuse repository and persistent website \citep{RopentreeTutorials, RopentreeTutorialsWebsite}.
The tutorial is available for the users to go back to any time they need it,
and to be passed on to other users (Figure \ref{fig:second}).

@@

We created a main version of the tutorial that is stable. Any updates to the tutorial are
published as new versions, or tutorials for new workshops \citep{CARPENTRIES}
 Versions presented at workshops are a copy from the original repository.
They represent a temporally stable snapshot of functions and workflows presented
during a workshop.

@@

Response form the community has been invaluable in gauging success of the tutorials.
Senior researchers often comment on the usefulness
of the tutorials for their research, as well as supporting students in using the
R packages with less help from them as PIs.

Making accessible reproducible workflows has several advantages:

- save explanation/training time when analyses are run again by students and collaborators;
- save research time for yourself when analyses are run again with more data, a
  different dataset, a different organism or biological model;
- scientific efforts can build off of each other.

@@

Some universities have been incorporating the subject in their classes (see
\cite{uwlibraries2022, nigms2022}).
The focus of these resources has been for students to develop skills to document their work.
The principles identified and outlined here can be used to set learning goals and
outcomes on new reproducibility syllabi.

The principles to create tutorials described here facilitate adoption of software
 and analysis workflows among researchers at different academic levels, from undergrads
  to established researchers.
It will also help close the gap between students that had access to computational
resources (and computational training) from an early age and students that did not.
Late access to computational resources and training can occur due to lack of
economic resources, often occurring in households from underrepresented communities
and minorities \citep{REFERENCE-NEEDED}.
It can also be due to gender-biased parental and community pressures,
in which male individuals are more often encouraged to perform activities related to computers,
while female individuals are discouraged \citep{REFERENCE-NEEDED}.
% How to balance software acceptance VS. adoption?
These principles can be used to improve not only reproducibility practices,
 but also software adoption in the natural sciences.
%% Discuss: why address accessibility and not other aspects of reproducibility?
